---
title: "Lab18_CLOM"
author: "Celia L Ochoa Medina"
format: html
---

Pertussis and the CMI-PB project

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L,2024L),
  cases = c(107473,164191,165418,152003,
                                   202210,181411,161799,197371,
                                   166914,172559,215343,179135,265269,
                                   180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,
                                   120718,68687,45030,37129,60886,
                                   62786,31732,28295,32148,40005,
                                   14809,11468,17749,17135,13005,6799,
                                   7717,9718,4810,3285,4249,3036,
                                   3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,
                                   3589,4195,2823,3450,4157,4570,
                                   2719,4083,6586,4617,5137,7796,6564,
                                   7405,7298,7867,7580,9771,11647,
                                   25827,25616,15632,10454,13278,
                                   16858,27550,18719,48277,28639,32971,
                                   20762,17972,18975,15609,18617,
                                   6124,2116,3044, 35435)
)
```

```{r}
head(cdc)
```
Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(x=year, y=cases) +
  geom_line() +
  labs(title = "Pertussis Cases by Year (1922-2019)")
```

Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(x=year, y=cases) +
  geom_line() +
  geom_vline(xintercept = 1942, col="blue")+
  geom_vline(xintercept = 1996, col = "red")+
  geom_vline(xintercept = 2020, col="purple") +
  geom_vline(xintercept = 2004, col="grey") +
  labs(title = "Pertussis Cases by Year (1922-2019)")

```

There are high case prior to 1946, before wP vaccine. Then there is a rapid decrease in case numbers through the 1970s and on to 2004 when the first widespread outbreak occurs again. 

This suggests that the aP vaccine induced immunity wanes faster than the older xP vaccine. 

Enter the CMI-PB Project 

>Computational Models of Immunity Pertussis Boost

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject)
```

>Q4 How many aP and wP subject are there?

```{r}
table(subject$infancy_vac)
```

>Q How many individuals "subjects" are there in this dataset?

```{r}
nrow(subject)
```

>Q5 How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```


> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table (subject$race,subject$biological_sex)
```

>Q. Does this look to be representative of the US population at large?

No. 


Let's read some more CMI-PB data 

```{r}
library(lubridate)
today()
today() - ymd("2000-01-01")
round(time_length( today() - ymd("2000-01-01"),  "years"),2)
```
> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}

# Use todays date to calculate age in days
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")

```

>Q8. Determine the age of all individuals at time of boost? 

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")

```




```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)

```



```{r}
head(specimen)
```
```{r}
head(ab_titer)
```

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
dim(meta)
```
>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
ab_data <- inner_join(meta, ab_titer,)

dim(ab_data)
```
>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(ab_data$isotype)
```

>Q. How many differnet anigens?

```{r}
table(ab_data$antigen)
```


> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

>Q. Let's look at a box plot of antigen levels over the whole dataset?


```{r}
igg <- ab_data %>% filter(isotype == "IgG")
head(igg)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(ab_data)+
  aes(MFI, antigen)+
  geom_boxplot()
  
```

>Q. Break this plot down by aP or wP.
Can have different color for each 

```{r}
ggplot(ab_data)+
  aes(MFI, antigen, col=infancy_vac)+
  geom_boxplot()
```


```{r}
ggplot(ab_data)+
  aes(MFI, antigen)+
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

Another version of this plot adding infancy_vac to the faceting:


```{r}
ggplot(ab_data)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```



```{r}
igg <- ab_data |> 
  filter(isotype =="IgG")
```



```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```


```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot() +
  facet_wrap(~visit)
```



PT(petussis Toxin) and igg levels over time

```{r}
table(ab_data$dataset)
```
Filter to focus on one antigen, PT, and IgG levels 

```{r}
pt_igg <- ab_data |> 
  filter(isotype=="IgG", antigen=="PT", dataset == "2021_dataset")
```

A plot of `actual days relative to boost ` vs `MFI _normalised`
```{r}
ggplot(pt_igg)+
  aes(actual_day_relative_to_boost, MFI_normalised, col=infancy_vac, group = subject_id)+
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = 14) +
  geom_vline(xintercept = 0)
  labs(titile = "IgG PT Levels", 
       xlab = "Days relative to Boost") +
    geom_smooth(method= "loess", se = T)
  
```


>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
ggplot(pt_igg)+
  aes(MFI_normalised, col=infancy_vac)+
  geom_boxplot()
  
```










